### 2025-09-18 paper/sec/introduction.tex
Rehashing intro.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25} approaches the
transparency problem from this direction, through software infrastructure which transparently connections
charts and figures to their data sources. \rpnote{Contrast with LLM-based approach -- trusted infrastructure.}
This enables visual outputs to provide interactive \emph{provenance queries}, user interactions (e.g. mousing
over a visual element) that allow readers to explore how visual features relate to data. The value proposition
of these approaches is that the interactions are provided automatically via trusted infrastructure, namely an
underlying programming language (say query language or general-purpose language) which tracks dependencies as
data flows through a computation. However, these approaches are limited to outputs computed from data, such as
visualisations.

In this paper, we extend interactive provenance queries to the natural language that accompanies the charts
and figures. We believe this is an important step because in most articles, charts and figures usually only
play a supporting role. The main claims are usually presented in natural language, language which also makes
quantitative statements and contains substantial implicit analytic content. In principle transparent
programming languages can also be used to create natural language which is bidirectly linked to data, by
computing key fragments of natural language from data. The following example, adapted from \cite{zhang18} via
\cite{moosavi21}, shows how the language presented in this paper can be used to achieve this \rpnote{Need to
illustrate counterfactual/generalisability}:

>>>>

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25} approaches the
transparency problem from this direction, through software infrastructure which transparently connections
charts and figures to their data sources. \rpnote{Contrast with LLM-based approach -- trusted infrastructure.}
This enables visual outputs to provide interactive \emph{provenance queries}, user interactions (e.g. mousing
over a visual element) that allow readers to explore how visual features relate to data. The value proposition
of these approaches is that the interactions are provided automatically via trusted infrastructure, namely an
underlying programming language (say query language or general-purpose language) which tracks dependencies as
data flows through a computation. However, these approaches are limited to outputs computed from data, which
usually precludes natural language, despite the fact that the main claims of a paper are usually made using
natural language, with charts and figures only playing a supporting role.

In this paper, we make the case for combining these two approaches, leveraging the ability of LLMs to
understand complex technical natural language and to synthesise programs that express the analyses that
underwrite those claims, and extend interactive provenance queries to the natural language that accompanies
the charts and figures.  ..language which also makes quantitative statements and contains substantial implicit
analytic content. In principle transparent programming languages can also be used to create natural language
which is bidirectly linked to data, by computing key fragments of natural language from data. The following
example, adapted from \cite{zhang18} via \cite{moosavi21}, shows how the language presented in this paper can
be used to achieve this \rpnote{Need to illustrate counterfactual/generalisability}:
